Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex

\begin{figure}
\ADComment{Tyler/Hugues: please insert graph traversal example.}
\TSComment{Hows this?}

\begin{lstlisting}
kernel graph_traversal(global graph *g, 
                       global nodes *n0, global nodes *n1) {

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups, stating~\cite{...}: \emph{``A conforming implementation
  may choose to serialize the workgroups so a correct algorithm cannot
  assume that workgroups will execute in parallel.  There is no safe
  and portable way to synchronize across the independent execution of
  workgroups since once in the work-pool, they can execute in any
  order.''}  OpenCL is not alone in this regard: CUDA similarly
provides no guarantees~\cite{...}, and HSA guarantees only that
\ADComment{Tyler, can you state the HSA guarantee?}~\cite{...}.

@@

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier()$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented following any inter-workgroup barrier
design, e.g.~\cite{...}, and the atomic operations provided by the
OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{...}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

@@

@@

@@

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel.  Although $\getnumgroups()$ has
atomic semantics, and so can be called at any time to yield a
well-defined value, the value returned by two successive calls may
differ if workgroups have executed $\offerkill$, $\offerfork$ or
$\resizingglobalbarrier$ in-between.  The programmer must therefore
write their cooperative kernel in a manner that is robust to changes
in the value returned by $\getnumgroups()$.

@@

This guarantee can be exploited by algorithms that perform strided
data processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?}

>>>>

\begin{figure}

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups \TSComment{This should explicitly say \emph{workgroups
    executing the same kernel}, otherwise HSA has some weirdness},
stating~\cite{...}: \emph{``A conforming implementation may choose to
  serialize the workgroups so a correct algorithm cannot assume that
  workgroups will execute in parallel.  There is no safe and portable
  way to synchronize across the independent execution of workgroups
  since once in the work-pool, they can execute in any order.''}
OpenCL is not alone in this regard: CUDA similarly provides no
guarantees~\cite{...}.

HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... A
  comes after B in work-group flattened ID order}. However, this
one-way communication is not sufficient so support either the
work-stealing or graph application examples of
Figure~\ref{fig:graphsearch} and \ref{fig:workstealing}. As described
above, both require \emph{symmetric} communication between
threads. \ADComment{Tyler, can you state the HSA guarantee?},
\TSComment{How is this?}~\cite{...}.

@@

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier()$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented following any inter-workgroup barrier
design, \TSComment{This isn't exactly true. For cooperative kernels,
  the barriers must be able to handle a growing or shrinking number of
  workgroups.} e.g.~\cite{...}, and the atomic operations provided by
the OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{...}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

@@

\begin{figure}

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {

(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;

(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();

    level++;
    swap(in_nodes, out_nodes);
    reset(out_nodes);
(*@\label{line:cgraph:resizing}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
\caption{Cooperative kernel port for Figure~\ref{fig:graphsearch}}\label{fig:cgraphsearch}
\end{figure}

@@

In practice, we found that transforming traditional blocking GPU
kernels into cooperative kernels was straightforward. For example,
Figure~\ref{fig:cgraphsearch} shows the graph traversal example of
Figure~\ref{fig:graphsearch} transformed into a cooperative kernel.
On line~\ref{line:cgraph:resizing}, we change the original barrier
into a resizing barrier. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barrier
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): {\tt level} must be restored so that
new workgroups know which frontier they are processing; {\tt
  in\_nodes} and {\tt out\_nodes} must be restored so that new
workgroups know which of the node arrays are input and output. Lastly,
the static work distribution of the original kernel is no longer valid
in a cooperative kernel. This is because the stride (which is based on
$M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}).

\TSComment{Can Hugues talk about work stealing here?}

@@

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel \TSComment{Neither is {\tt
    get\_global\_size}, is that worth mentioning?}.  Although
$\getnumgroups()$ has atomic semantics, and so can be called at any
time to yield a well-defined value, the value returned by two
successive calls may differ if workgroups have executed $\offerkill$,
$\offerfork$ or $\resizingglobalbarrier$ in-between.  The programmer
must therefore write their cooperative kernel in a manner that is
robust to changes in the value returned by $\getnumgroups()$.

@@

This guarantee can be exploited by algorithms that perform strided
data processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?} \TSComment{Done,
its in figure~\ref{fig:cgraphsearch}}
